Drawing Trees and Animating Tree Changes

### 2010-05-14 paper/paper.tex
Possibly the one I am submitting in an hour.

\begin{abstract} Dynamic trees appear in many areas of computer science,
including, but not limited to, Lisp code, data structures, tree automata, and
computer algorithms.  It would be useful to be able to easily draw trees and
animate them changing over time.  We begin by articulating a design for tree
visualization and animation that is more general than previous methodologies.
Our design is enabled by the observation that, for trees which are small enough
to be displayed on a screen, computers are now fast enough to eliminate many
worries about efficiency.  This allows us to make a library which emphasizes
simplicity of use over efficiency of execution.  We demonstrate our library
using several visualizations.\end{abstract}

\section{Introduction}
Trees, and especially trees which change over time, are fundamental objects in
computer science.  They occur in many disparate contexts, and while the uses of
these trees may be different, there exist commonalities which we can use to
build a tree visualization system which might be useful in many different
contexts.  In this paper we articulate the design principles of a highly usable
tree visualization system, present one such system that we have built in Java
and made freely available, and then demonstrate our system's utility with
several sample visualizations.

@@

Computing and math eductors have a long history of building visualization
systems for demonstration, on the grounds that if students can view these
systems in action, then the students will understand these system.
Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
visualizations are useless\cite{hds}.  A good visualization seems intuitively useful, but
it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.  The one bright spot in this was in interactivity and construction.  The one activity that actually improved student understanding was not the viewing of a visualization, but instead its construction.

@@

For educational visualization to be effective, students must write their own
visualizations.  Therefore, whenever possible, we design our system to be as
simple, learnable, and robust as possible.  This will allow more students to
succeed in the creation of visualizations by attemping to make the graphics
aspect of the programming task as simple as possible.

@@

Computational processes are often both sophisticated and beautiful, but their
beauty can be lost in the nitty-gritty of simply getting a given algorithm to
work.  Using a visualization system could, potentially, remind students of the
beauty of the structures they are working with and also serve as an effective
advertisement to non-majors of the intricacy and interestingness available in
the computer science and mathematics majors.

To acheive this goal of beauty, we will pay attention to the aesthetic aspects
of our visualization as well as maintaining correctness.  In particular, we
should make sure that transitions are smooth and not jarring, and that our
layouts are as aesthetically pleasing as possible.  This requirement is
difficult to quantify, but, cetus paribus, a beautiful visualization is always
better than an ugly one.

@@

Let us begin our design by formally defining the objects under consideration.
A \newword{tree} $T$ is an undirected connected acyclic graph with vertex set
$V$ and edge set $E \subseteq \{V \times V\}$.  A \newword{rooted tree} is a
tree where a particular vertex $r$ is designated the as the \newword{root} of
the tree, which allows us to define the \newword{depth} of a vertex $v$ as the
length of the shortest path between $v$ and $r$. Every edge connects two
vertices $(u,v)$, and we will assume, again wlog, that the depth of $u$ is one
less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
$v$ and $v$ the \newword{child} of $u$.  The \newword{children} of $v$ are the
set of vertices which have $v$ as their parent.  A \newword{rooted ordered
tree} is a rooted tree with a partial-ordering on the vertex set $V$ such that
for all vertices $\{u,v, \ldots\}$ with the same parent, either $u < v$ or $v <
u$.

Without loss of generalization, we restrict our visualization system to rooted,
ordered trees.  Any tree can be converted into a rooted, ordered tree by
selecting a root and imposing a partial-ordering on the vertices.  In practice,
however, we found that every domain we considered had a naturally occuring root
and ordering, and that the imposition of an arbitrary root or an arbitrary
ordering was unnecessary.  From here forward, when we say tree, we will mean a
rooted ordered tree.

@@

Starting with this initial layout, we use force-directed layout, but we place
constraints on how the vertices can move in response to the forces applied.
Most importantly, we only allow vertices to left left-right, and never
up-down.  The initial vertical position of a vertex is also its final
position.  Similarly, a vertex may never swap rank with another vertex on the
same level --- no matter how strong the forces involved, the initial ordering
of a row must also be the final ordering of a row.  Metaphorically, one can
think of treating the vertices as train cars on a horizontal rail.  The cars
can only move back and forth in one direction, and they may never move through
each other.

In a large graph, this layout may take some time to settle down (on the order
of seconds).  We have seen positive effects in displaying the layout process to
the user.  In particular, multiple users have commented how, by showing the
layout process settling the tree into the final layout, the graph seems more
``alive''.

@@

Either on every tree transformation or when requested by the visualization
programmer, we relayout (off-screen) the entire tree according to our layout
algorithm and then we animate the transition from the old tree to the new
tree.  Or animation procedure is extremely basic, but the simplicity allows for simplicity of use.

Our animation procedure has six steps.  First, we lay out, off-screen, the new
tree.  Second, we zoom out (if required) so that the viewable area is the size
of the maximum of the new tree and old tree.  Third, we fade out any of the
vertices and edges which were in the old tree, but have been removed in the new
tree. Fourth, we animate the movement of all vertices which are present in both
trees from their old positions to their new postions.  Fifth, we fade in any
new nodes which are present only in the new tree.  Sixth, we zoom in (if
required) so that the new tree fills the viewing area.

Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This allows artistic flexibility when creating a new visualization, but allows visualization programmers to err on the side of more visualization rather than less.

@@

Our library\cite{ourlibrary} is written in Java, and users of the library need only use two of
the classes it contains: the \java{Tree} class, and the \java{TreeDisplay}
class.  Our primary design principle in building this system was ease of usage
by another programmer.  Therefore, our two classes have very few methods, and
those few methods are very simple to use.  Prior efforts focused a lot of work
on making specific animations efficient, but we have instead focused our
efforts on ease of use.

@@

The class \java{TreeDisplay} needs to only be created once.  A
\java{TreeDisplay} is an AWT component, and may be used as such in any Java
code.  A \java{Tree} may be created at any time, and to display a tree, use the
\java{setTree} method of the \java{TreeDisplay} class.  Each \java{Tree} has
methods to manipulate both the data displayed at the root of that particular
subtree, as well as to manage the children of the root note of the subtree,
which are, naturally, trees themselves.

Each \java{Tree} has no ties to any algorithm.  The user must program any manipulation algorithm themselves.  If the \java{Tree} is currently being displayed, then any tree operation will be animated by default.

@@

The tree automaton we visualize accepts all trees which represent logical tautologies made out of the logical operations ``and'', ``or'', and ``not'', and the literals 0 and 1.  Our rewrite rules for the tree are shown in Figure~\ref{fig:rules}, and the set of accepted trees is a tree with a single vertex named ``true''.

@@

The progressive execution of the tree automaton consists of recognizing some
subtree which conforms to the left hand side of a rule and then replacing that
subtree with the right side of the rule.  Our visualization library alowed us
to implement a tree automation and its attendant visualization using
approximately 200 lines of Java code.  The size is mentioned because a project of that size is small enough to be appropriate for a class assignment, which is a positive sign that we are meeting the requirements of Section~\ref{educ}.

Our visualization of this automaton analyzing the logical expression \\
~\\
($\lnot$ (($\lnot$ ((1 $\lor$ 1) $\lor$ ($\lnot$ 1))) $\land$ ((($\lnot$ 1) $\land$ (0 $\land$ 0)) $\lor$ ((1 $\lor$ 1) $\lor$ (1 $\lor$ 0))))) \\
$\land$ ($\lnot$ (((($\lnot$ 1) $\lor$ (0 $\land$ 1)) $\land$ ((0 $\lor$ 0) $\land$ (1 $\lor$ 1))) $\land$ (((1 $\land$ 0) $\lor$ (0 $\land$ 1)) $\lor$ ((1 $\land$ 1) $\lor$ (0 $\lor$ 0)))))\\
~\\
may be seen in flipbook form in Figure~\ref{fig:auto}.  Interesting things of note include a few frames of the animation where a vertex is in the process of ``fading out'', as well as the successive zooming as the tree gets smaller and smaller.  Note that our final frame indicates that the tree automaton has accepted the input, and that therefore the logical expression was a tautology.

@@

There exist multiple virtual machines which model the execution of lisp code in equivalent manners, here we used one based on string substitution.  The interested reader is referred to {\it Abstract computing Machines} by Kluge\cite{acm}.  This particular visualization took more code than the tree automaton, but it includes the implementation of a turing-complete lisp compiler and virtual machine, and so may be forgiven for the extra length.  The total length of all the code required was one the order of 450 lines of Java, and so is suitable for a long project or a perhaps a group project in a programming languages course.

The results of our visualization may be found in the animation flipbook shown
in Figure~\ref{lispanim}. In the flipbook, one can see many examples of
vertices caught in the middle of an animation.  There are multiple examples of
vertices fading out or caught in the middle of moving around the screen.

@@

As well as the execution of the evaluation of the quadratic formula, we have also produced visualizations of lisp code calculating the factorial function and the Fibonacci sequence.  These examples, while interesting, are mostly interesting as examples of the $\lambda$-calculus and teaching aids in a programming languages course.  The animation proceeds much like the the flipbook shown, but the trees involved are much larger and more inctricate.

\subsection{Other example visualizations}

Also included with our library are visualizations of self-adjusting binary search trees (AVL Trees) and of the growth of a random tree.  The random tree is particularly interesting, because it was requested by an audience member during a presentation and was then implemented in the next 30 minutes.  It is hoped that this short distance from request to implementation can serve as an example of the power of both our library and especially our design approach.

>>>>

\begin{abstract} Dynamic trees appear in many areas of computer science,
including, but not limited to, Lisp code, data structures, tree automata, and
computer algorithms.  It would be useful to be able to easily draw trees and
animate them changing over time.  We begin by articulating a design for tree
visualization and animation that is more general than previous methodologies.
Our design is enabled by the observation that, for trees which are small enough
to be displayed on a screen, computers are now fast enough to eliminate many
worries about efficiency.  This observation allows us to make a library which
emphasizes simplicity of use over efficiency of execution.  We show how to use  our
library using several visualizations.\end{abstract}

\section{Introduction}
Trees, and especially trees which change over time, are fundamental objects in
computer science.  They occur in many disparate contexts, and while the uses of
these trees may be different, there exist commonalities which we can use to
build a tree visualization system which might be useful in many different
contexts.  We will articulate the design principles of a highly usable
tree visualization system, present one such system that we have built in Java
and made freely available, and then show our system's utility with
several sample visualizations.

@@

Computing and math educators have a long history of building visualization
systems for demonstration.  The intuition in many case was that if students can view these
systems in action, then the students will understand these system.
Unfortunately for those goals, Hundhausen, Douglas, and Stasko found that, in practice, almost all
visualizations are useless\cite{hds}.  A good visualization seems intuitively useful, but
it turns out that if a student does not understand the system being visualized, they will not understand the visualization either.  The one bright spot in this was in interactivity and construction.  The one activity that improved student understanding was not the viewing of a visualization, but instead its construction.

@@

For effective educational visualization, students must write their own
visualizations.  Therefore, whenever possible, we design our system to be as
simple, learnable, and robust as possible.  This will allow more students to
succeed in the creation of visualizations by attempting to make the graphics
aspect of the programming task as simple as possible.

@@

Computational processes are often both sophisticated and beautiful, but their
beauty can be lost in the nitty-gritty of simply making a given algorithm
work.  Using a visualization system could, potentially, remind students of the
beauty of the structures they are working with and also serve as an effective
advertisement to non-majors of the intricacy and interestingness available in
the computer science and mathematics majors.

To achieve this goal of beauty, we will pay attention to the aesthetic aspects
of our visualization as well as maintaining correctness.  In particular, we
should ensure that transitions are smooth and not jarring, and that our
layouts are as aesthetically pleasing as possible.  This requirement is
difficult to quantify, but, cetus paribus, a beautiful visualization is always
better than an ugly one.

@@

Let us begin our design by formally defining the objects under consideration.
A \newword{tree} $T$ is an undirected connected acyclic graph with vertex set
$V$ and edge set $E \subseteq \{V \times V\}$.  A \newword{rooted tree} is a
tree where a particular vertex $r$ is designated the as the \newword{root} of
the tree, which allows us to define the \newword{depth} of a vertex $v$ as the
length of the shortest path between $v$ and $r$. Every edge connects two
vertices $(u,v)$, and we will assume, again without loss of generality, that the depth of $u$ is one
less than the depth of $v$.  In that case, we call $u$ the \newword{parent} of
$v$ and $v$ the \newword{child} of $u$.  The \newword{children} of $v$ are the
set of vertices which have $v$ as their parent.  A \newword{rooted ordered
tree} is a rooted tree with a partial-ordering on the vertex set $V$ such that
for all vertices $\{u,v, \ldots\}$ with the same parent, either $u < v$ or $v <
u$.

Without loss of generalization, we restrict our visualization system to rooted,
ordered trees.  Any tree can be converted into a rooted, ordered tree by
selecting a root and imposing a partial-ordering on the vertices.  In practice,
however, we found that every domain we considered had a naturally occurring root
and ordering, and that the imposition of an arbitrary root or an arbitrary
ordering was unnecessary.  From here forward, when we say tree, we will mean a
rooted ordered tree.

@@

Starting with this initial layout, we use force-directed layout, but we place
constraints on how the vertices can move in response to the forces applied.
Most importantly, we only allow vertices to left left-right, and never
up-down.  The initial vertical position of a vertex is also its final
position.  Similarly, a vertex may never swap rank with another vertex on the
same level. No matter how strong the forces involved, the initial ordering
of a row must also be the final ordering of a row.  Metaphorically, one can
think of treating the vertices as train cars on a horizontal rail.  The cars
can only move back and forth in one direction, and they may never move through
each other.

In a large graph, this layout may take some time to settle down (on the order
of seconds).  We have seen positive effects in displaying the layout process to
the user.  In particular, multiple users have commented how, by showing the
layout process settling the tree into the final layout, the graph seemed more
``alive'' on the screen.

@@

Either on every tree transformation or when requested by the visualization
programmer, we re-layout (off-screen) the entire tree according to our layout
algorithm and then we animate the transition from the old tree to the new
tree.  Or animation procedure is extremely basic, but the simplicity allows for simplicity of use.

Our animation procedure has six steps.  First, we lay out, off-screen, the new
tree.  Second, we zoom out (if required) so that the viewable area is the size
of the maximum of the new tree and old tree.  Third, we fade out any of the
vertices and edges which were in the old tree, but have been removed in the new
tree. Fourth, we animate the movement of all vertices which are present in both
trees from their old positions to their new positions.  Fifth, we fade in any
new nodes which are present only in the new tree.  Sixth, we zoom in (if
required) so that the new tree fills the viewing area.

Multiple changes can be either combined into a single animation step, or each step may be animated individually, which is the default  This choice allows artistic flexibility when creating a new visualization, but encourages visualization programmers to err on the side of more visualization rather than less.

@@

Our library\cite{ourlibrary} is written in Java, and users of the library need only use two of
the classes it contains: the \java{Tree} class, and the \java{TreeDisplay}
class.  Our primary design principle in building this system was ease of usage
by another programmer.  Therefore, our two classes have very few methods, and
those few methods are very simple to use.  Prior efforts focused 
on making specific animations efficient, but we have instead focused our
efforts on ease of use.

@@

The class \java{TreeDisplay} needs to only be created once.  A
\java{TreeDisplay} is an AWT component, and may be used as such in any Java
code.  A \java{Tree} may be created at any time, and to display a tree, use the
\java{setTree} method of the \java{TreeDisplay} class.  Each \java{Tree} has
methods to manipulate both the data displayed at the root of that particular
subtree, as well as methods to manage the children of the subtree.
Those children are, naturally, trees themselves.

Each \java{Tree} has no ties to any algorithm.  The user must program any manipulation algorithm themselves.  If the \java{Tree} is being actively displayed, then any tree operation will be animated by default.

@@

The tree automaton we visualize accepts all trees which represent logical tautologies made out of the logical operations ``and'', ``or'', and ``not'', and the literals 0 and 1.  Our rewrite rules for the tree may be found in Figure~\ref{fig:rules}, and the set of accepted trees is a tree with a single vertex named ``true'' and no edges.

@@

The progressive execution of the tree automaton consists of recognizing some
subtree which conforms to the left hand side of a rule and then replacing that
subtree with the right side of the rule.  Our visualization library allowed us
to implement a tree automation and its attendant visualization using
approximately 200 lines of Java code.  The size is mentioned because a project of that size is small enough to be appropriate for a class assignment, which is a positive sign that we are meeting the requirements of Section~\ref{educ}.

Our visualization of this automaton analyzing the logical expression \\
~\\
($\lnot$ (($\lnot$ ((1 $\lor$ 1) $\lor$ ($\lnot$ 1))) $\land$ ((($\lnot$ 1) $\land$ (0 $\land$ 0)) $\lor$ ((1 $\lor$ 1) $\lor$ (1 $\lor$ 0))))) \\
$\land$ ($\lnot$ (((($\lnot$ 1) $\lor$ (0 $\land$ 1)) $\land$ ((0 $\lor$ 0) $\land$ (1 $\lor$ 1))) $\land$ (((1 $\land$ 0) $\lor$ (0 $\land$ 1)) $\lor$ ((1 $\land$ 1) $\lor$ (0 $\lor$ 0)))))\\
~\\
may be seen in flipbook form in Figure~\ref{fig:auto}.  Interesting things of note include a few frames of the animation where a vertex is in the process of ``fading out'', as well as the successive zooming as the tree gets smaller and smaller.  The final frame of the flipbook indicates that the tree automaton has accepted the input, and that therefore the logical expression was a tautology.

@@

There exist multiple virtual machines which model the execution of lisp code in equivalent manners, here we used one based on string substitution.  The interested reader is referred to {\it Abstract computing Machines} by Kluge\cite{acm}.  This particular visualization took more code than the tree automaton, but it includes the implementation of a Turing-complete lisp compiler and virtual machine, and so may be forgiven for the extra length.  The total length of all the code required was one the order of 450 lines of Java, and so is suitable for a long project or a perhaps a group project in a programming languages course.

The results of our visualization may be found in the animation flipbook shown
in Figure~\ref{lispanim}. In the flipbook, one can see many examples of
vertices caught in the middle of an animation, either fading out or in the process of moving about the screen.

@@

As well as the execution of the evaluation of the quadratic formula, we have
also produced visualizations of lisp code calculating the factorial function
and the Fibonacci sequence.  These examples, while interesting, are mostly
interesting as examples of the $\lambda$-calculus and teaching aids in a
programming languages course.  The animation proceeds much like the flipbook
shown, but the trees involved are much larger and more intricate.

\subsection{Other Example Visualizations}

Also included with our library are visualizations of self-adjusting binary
search trees (AVL Trees) and of the growth of a random tree.  The random tree
is particularly interesting, because it was requested by an audience member
during a presentation and was then implemented in the next half hour.  We hope
that this short distance from request to implementation can serve as an example
of the power of both our library and especially our design approach.
